Continous time Markov chain models; \\ extension, simulation and visualisation

### 2014-11-17 chapters/chapter01.tex
labels

%bayesian
The advent of powerful and increasingly more attainable computing power has been instrumental to the adoption of statistical methods based on likelihood estimation and the closely related Bayesian methods.
Bayesian statistical thinking is not a new idea and stems back to the 18th century work of reverend Thomas Bayes, a statistician, philosopher and theologian.

>>>>

%bayesian
The advent of powerful and increasingly more attainable computing power has been instrumental to the adoption of statistical methods based on likelihood estimation and the closely related Bayesian methods.
Bayesian statistical thinking is not a new idea and stems back to the XVII century works of reverend Thomas Bayes, a statistician, philosopher and theologian.

\clearpage

### 2014-11-17 chapters/chapter02.tex
labels

From the rate matrix $\mathbf{Q}$, a stochastic matrix $\mathbf{P}$ is computed over time $t \geq 0$ via matrix exponentiation $\mathbf{P}(t)=
\exp(t\mathbf{Q})
=\underset{j=0}{\overset{\infty}{\sum}}
(t\mathbf{Q})^{j} / j!$. For an overview of methods to numerically approximate a matrix exponential, we refer to \citet{Moler1978}.
%
Drawing realizations with probabilities defined by $\mathbf{P}$ gives rise to 
a stochastic process $\left\{ X(t):\; t\geq0\right\}$ satisfying the Markov property, such that for every $n\geq 0$, given the time points $0\leq t_{0}\leq t_{1}<\ldots<t_{n}\leq t_{n+1}$ and discrete states $i_{0},i_{1}, \ldots, i_{n},i_{n+1}$ it holds that $P\left\{ X(t_{n+1})=i_{n+1}\mid X(t_{n})=i_{n},\ldots, X(t_{0})=i_{0}\right\} =P\left\{ X(t_{n+1})=i_{n+1}\mid X(t_{n})=i_{n}\right\} $. 
%
In general, one refers to the elements of $\mathbf{P}$ as finite-time transition probabilities between the $K$ discrete state-space elements. Let us denote a transition probability between two states $i$ and $j$ over time $u$ to $t+u$ by

@@

In the phylogenetic setting, researchers often further constrain these processes 
to be time-homogeneous and time-reversible. Time-homogeneity mandates that  transition probabilities depend only on the difference $t$ between times $u$ and $t + u$,
\begin{equation}
p_{ij}\left(u,t+u\right)= p_{ij}\left(0,t\right) \equiv p_{ij}\left(t\right) .
\end{equation}
Time-reversible CTMCs satisfy detailed balance, such that
$
\pi_{i}p_{ij}(t)=\pi_{j}p_{ji}(t)
%,
$
for all $i$, $j$ and $t$, where $\pi_j = p_{ij}(\infty)$ for all $j$ return the stationary distribution of the CTMC and are independent of starting state $i$.  Finally, common practice in phylogenetics reparametrizes the elements of $\mathbf{Q}$ into relative rates through the constraint $\sum_i \pi_i q_{ii} = -1$ and then, for studies involving phylogenies set in calendar time, multiples $\mathbf{Q}$ by a rate scalar $r$ to form the argument to $\mathbf{P}(t)$.  In this case, we define
\begin{equation}
p_{ij}(u,t+u,r) 
=
\left\{
\exp(
r  t  \mathbf{Q})
\right\}_{ij}
,
\label{eq:makeprobs}
\end{equation}
where $\{ \cdot \}_{ij}$ extracts the $ij$-th element.

>>>>

From the rate matrix $\mathbf{Q}$, a stochastic matrix $\mathbf{P}$ is computed over time $t \geq 0$ via matrix exponentiation $\mathbf{P}(t)=
\exp(t\mathbf{Q})
=\underset{j=0}{\overset{\infty}{\sum}}
(t\mathbf{Q})^{j} / j!$. For an overview of methods to numerically approximate a matrix exponential, we refer to \citet{Moler1978}.
%
Drawing realizations with probabilities defined by $\mathbf{P}$ gives rise to a stochastic process $\left\{ X(t):\; t\geq0\right\}$ satisfying the Markov property, such that for every $n\geq 0$, given the time points $0\leq t_{0}\leq t_{1}<\ldots<t_{n}\leq t_{n+1}$ and discrete states $i_{0},i_{1}, \ldots, i_{n},i_{n+1}$ it holds that %\newline
%
$P\left\{ X(t_{n+1})=i_{n+1}\mid X(t_{n})=i_{n},\ldots, X(t_{0})=i_{0}\right\} =P\left\{ X(t_{n+1})=i_{n+1}\mid X(t_{n})=i_{n}\right\} $. 
%
In general, one refers to the elements of $\mathbf{P}$ as finite-time transition probabilities between the $K$ discrete state-space elements. Let us denote a transition probability between two states $i$ and $j$ over time $u$ to $t+u$ by

@@

In the phylogenetic setting, researchers often further constrain these processes 
to be time-homogeneous and time-reversible. Time-homogeneity mandates that  transition probabilities depend only on the difference $t$ between times $u$ and $t + u$,

\begin{equation}
p_{ij}\left(u,t+u\right)= p_{ij}\left(0,t\right) \equiv p_{ij}\left(t\right) .
\end{equation}

Time-reversible CTMCs satisfy detailed balance, such that $\pi_{i}p_{ij}(t)=\pi_{j}p_{ji}(t)$ for all $i$, $j$ and $t$, where $\pi_j = p_{ij}(\infty)$ for all $j$ return the stationary distribution of the CTMC and are independent of starting state $i$.  
Finally, common practice in phylogenetics reparametrizes the elements of $\mathbf{Q}$ into relative rates through the constraint $\sum_i \pi_i q_{ii} = -1$ and then, for studies involving phylogenies set in calendar time, multiples $\mathbf{Q}$ by a rate scalar $r$ to form the argument to $\mathbf{P}(t)$. 
In this case, we define

\begin{equation}
p_{ij}(u,t+u,r) 
=
\left\{
\exp(
r  t  \mathbf{Q})
\right\}_{ij}
,
\label{eq:makeprobs}
\end{equation}

where $\{ \cdot \}_{ij}$ extracts the $ij$-th element.
